### 2021-07-08 rad-icfp-2021/complexity.tex
minor
several minor fixes

\begin{proof}[Notes]
    As $\directD{\rho}{\Gamma}{Y}$ is defined by induction on programs, it suffices to show locally 
that $\directD{\rho}{\Gamma}{Y}$ verifies the cheap gradient principle.

@@

In Figure~\ref{fig:optim} we present a of optimizations. 
These optimizations are all obviously valid according to the semantic model. 
As our language is purely functional, we can use these extra optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let-binding
and resembles SSA-code.

>>>>

\begin{proof}[Proof Sketch]
    As $\directD{\rho}{\Gamma}{Y}$ is defined by induction on programs, it suffices to show locally 
that $\directD{\rho}{\Gamma}{Y}$ verifies the cheap gradient principle.

@@

In Figure~\ref{fig:optim} we present a of optimizations. 
These optimizations are all obviously valid according to the semantic model. 
As our language is purely functional, we can use these extra optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let bindings 
and resembles SSA-code.

### 2021-07-08 rad-icfp-2021/figures/macro_diff.tex
several minor fixes

\begin{figure*}[t]
\begin{tabular}{|c|}
\hline
    \begin{tabular}{r c l}
    $\Dsynrevsymbol_{\rho}$(!A1,$\ldots$,An!) &=& !A1,$\ldots$,An,(A1x$\ldots$xAn)->!$\rho$\\ \\
    $\Dsynrevsymbol_{\rho}$(!var!$_{T;i}$) &=& $F$(!var$_{T;i}$, J$^T$var$_{T;i}$!) \\
    $\Dsynrevsymbol_{\rho}$(!op!$_{\Gamma;n}$) &=& $F$(!op$_{T;n}$,J$^T$op$_{T;,n}$!) \\ 
    $\Dsynrevsymbol_{\rho}$(!pair!$_{T;A\times B}$) &=& $F$(!pair$_{T;A\times B}$, J$^T$pair$_{T;A\times B}$!) \\
    $\Dsynrevsymbol_{\rho}$(!proj!$_{T1;T2;T3}$) &=& $F$(!proj$_{T1;T2;T3}$,J$^T$proj$_{T1;T2;T3}$!) \\
    $\Dsynrevsymbol_{\rho}$(!e1!$\comp$!e2!) &=& $\Dsynrevsymbol_{\rho}$(!e1!); $\Dsynrevsymbol_{\rho}$(!e2!)\\ 
    $\Dsynrevsymbol_{\rho}$(!map2!$_{T;x,y.e}$) &=& $F$(!map2$_{T;x,y.e}$, J$^T$map2$_{T;x,y.e}$!) \\
    $\Dsynrevsymbol_{\rho}$(!reduce!$_{T;x,y.e1;e2}$) &=& $F$(!reduce$_{T;x,y.e1;e2}$, J$^T$reduce$_{T;x,y.e1;e2}$!) \\
    \end{tabular}\\
    where !$F$(A,B)$\defeq$ <proj$_{T;\underline{T}->\rho;[]} \comp$ A, proj$_{[];T;\underline{T}->\rho} \icomp$ (proj$_{T;\underline{T}->\rho;[]}$ $\comp$ B)>!\\\hline
    \end{tabular}
    \caption{Reverse-mode differentiation from Source UNF to Target UNF}
    \label{fig:diff_macro}    
\end{figure*}

>>>>

\begin{figure*}[t]
\begin{tabular}{|c|}
\hline
    \begin{tabular}{r c l}
    $\Dsynrevsymbol_{\rho}$(!A$_1$,$\ldots$,A$_n$!) &=& !A$_1$,$\ldots$,A$_n$,(A$_1\times\ldots\times$A$_n$)->$\rho$!\\ \\
    $\Dsynrevsymbol_{\rho}$(!var!$_{T;i}$) &=& $F$(!var$_{T;i}$, J$^T$var$_{T;i}$!) \\
    $\Dsynrevsymbol_{\rho}$(!op!$_{\Gamma;n}$) &=& $F$(!op$_{T;n}$,J$^T$op$_{T;n}$!) \\ 
    $\Dsynrevsymbol_{\rho}$(!pair!$_{T;A\times B}$) &=& $F$(!pair$_{T;A\times B}$, J$^T$pair$_{T;A\times B}$!) \\
    $\Dsynrevsymbol_{\rho}$(!proj!$_{T_1;T_2;T_3}$) &=& $F$(!proj$_{T_1;T_2;T_3}$,J$^T$proj$_{T1;T2;T3}$!) \\
    $\Dsynrevsymbol_{\rho}$(!e$_1\comp$e$_2$!) &=& $\Dsynrevsymbol_{\rho}$(!e$_1$!)$\comp$ $\Dsynrevsymbol_{\rho}$(!e$_2$!)\\ 
    $\Dsynrevsymbol_{\rho}$(!map2!$_{T;x,y.e}$) &=& $F$(!map2$_{T;x,y.e}$, J$^T$map2$_{T;x,y.e}$!) \\
    $\Dsynrevsymbol_{\rho}$(!reduce!$_{T;x,y.e_1;e_2}$) &=& $F$(!reduce$_{T;x,y.e_1;e_2}$, J$^T$reduce$_{T;x,y.e_1;e_2}$!) \\
    \end{tabular}\\
    where !$F$(A,B)$\defeq$ <proj$_{T;\underline{T}->\rho;[]} \comp$ A, proj$_{[];T;\underline{T}->\rho} \icomp$ (proj$_{T;\underline{T}->\rho;[]}$ $\comp$ B)>!\\\hline
    \end{tabular}
    \caption{Reverse-mode differentiation from Source UNF to Target UNF}
    \label{fig:diff_macro}    
\end{figure*}

### 2021-07-08 rad-icfp-2021/generalization.tex
several minor fixes

We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type $G$.
That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.

A ground type $G$ is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context $\Gamma=x1:G1,\ldots,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}Ai)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.

@@

\begin{tabular}{r c l}
    $\reals \otimes A$ & $\defeq$ & $A$ \\
    $(A1 \times \ldots \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times \ldots \times (An \otimes A)$ \\
    $\Array{A1}{n} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}{n}$
\end{tabular}

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldots\times An$ as expected.
For !map2!, we need a generalization of $*:\reals\to\reals$. 
If !e1: A!, then $\grad_{\{x\}}$!e1: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
$\widetilde{*}$ corresponds to matrix-matrix multiplication.
Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!, as can be seen in Figure~\ref{fig:lift_gen_arr}.

>>>>

We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type !G!.
That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.

A ground type !G! is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context !$\Gamma$=x$_1$:G$_1$,$\ldots$,x$_n$:G$_n$! is interpreted as $\bigoplus_{1\leq i\leq n}A_i$, where $A_i\defeq\seml$(!G$_i$!)$\semr$.
The denotation of the gradient of a term !$\Gamma \vdash$ e: G! at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}A_i)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}A_i\otimes A$.

@@

\begin{tabular}{r c l}
    $\reals \otimes A$ & $\defeq$ & $A$ \\
    $(A_1 \times \ldots \times A_n)\otimes A$ & $\defeq$ & $(A_1\otimes A) \times \ldots \times (A_n \otimes A)$ \\
    $\Array{A_1}{n} \otimes A$ & $\defeq$ & $\Array{A_1\otimes A}{n}$
\end{tabular}

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A_1\times\ldots\times A_n$ as expected.
For !map2!, we need a generalization of $*:\reals\to\reals$. 
If !e$_1$: A!, then $\grad_{\{x\}}$!e$_1$: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
$\widetilde{*}$ corresponds to matrix-matrix multiplication.
Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!, as can be seen in Figure~\ref{fig:lift_gen_arr}.

### 2021-07-08 rad-icfp-2021/unf.tex
several minor fixes

\begin{example}
    The term in Source UNF !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$! 
    intuitively represents a term \\
    !x$_1$:$\reals$, x$_2$:$\reals$, x$_3$:$\reals$ $\vdash$ let (x$_1$,x$_2$,x$_3$,x$_4$)=(x$_1$,x$_2$,x$_3$,cos(x$_3$)) in (x$_1$,x$_2$,<x$_3$,x$_4$>)!.
    The operator $\comp$ can more generally be understood as a let binding of a tuple, and primitives only act on the last parts of the tuple, always also returning their whole context.
\end{example}

@@

\begin{lemma}[Well typedness $\Dsynrevsymbol_{\rho}$]
    Let !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$! be a term in Source UNF. \\
    Then 
    % \begin{center}
        !A$_{1}$,$\ldots$,A$_{n}$,A$_{1}\times\ldots\times$ A$_{n}$->$\rho$ $\vdash$ $\Dsynrevsymbol_{\rho}$(e): B$_{1}$,$\ldots$,B$_{m}$,B$_{1}\times\ldots\times$ B$_{m}$->$\rho$!
    % \end{center}
\end{lemma}
\begin{proof}
    By induction on derivation of !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$!.
\end{proof}

@@

We first give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}, which we call $\UNFSymbol$.
As discussed above, $\UNFSymbol$ sends a constant or an operator to a primitive in Source UNF. 
This term in Source UNF carries the context of the original term in Source.
$\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
Because of this, we we're trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
When we need to pass the result of type !A! of !e$_{1}$! through !e$_{2}$! which does not take it as an input, we can transform !e$_{2}$! to accept and pass it.
It's the equivalent in UNF of a weakening.
This weakening is required when we transform a non-unary operator such as $+$ or !pair!, as we transform the arguments in order and we need to keep all of their results.
We write this new term $(\widetilde{e_{2}})_{A}$ and it defined by induction on !e$_{2}$! as follows. 
We assume given a type !A! that needs to be passed, and drop the index !A!.

@@

Next, the transformation from Target UNF to Target is presented in Figure~\ref{fig:unf_to_target}. 
We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of the ANF transformation \cite{sabry1993reasoning}. 
$\UNFSymbol^{-1}$ does not have as a simple typing property as $\UNFSymbol$, 
because it treats primitives with and without a $J$ differently and should be performed after $\Dsynrevsymbol_{\rho}$.

>>>>

\begin{example}
    The term in Source UNF !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$! 
    intuitively represents a term \\
    !x$_1$:$\reals$,x$_2$:$\reals$,x$_3$:$\reals$ $\vdash$ let (x$_1$,x$_2$,x$_3$,x$_4$)=(x$_1$,x$_2$,x$_3$,cos(x$_3$)) in (x$_1$,x$_2$,<x$_3$,x$_4$>)!.\\
    The operator $\comp$ can more generally be understood as a let binding of a tuple, and primitives only act on the last parts of the tuple, always also returning their whole context.
\end{example}

@@

\begin{example}
    Continuing with our previous example !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$!, we have

    \begin{tabular}{r c l}
        !$\Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}\comp$ pair$_{\reals,\reals; \reals\times\reals}$)!
        &=& !<> $\comp$! \\
        &&  !<> !
    \end{tabular}
\end{example}

\begin{lemma}[Well typedness $\Dsynrevsymbol_{\rho}$]
    Let !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$! be a term in Source UNF. \\
    Then  !A$_{1}$,$\ldots$,A$_{n}$,A$_{1}\times\ldots\times$ A$_{n}$->$\rho$ $\vdash$ $\Dsynrevsymbol_{\rho}$(e): B$_{1}$,$\ldots$,B$_{m}$,B$_{1}\times\ldots\times$ B$_{m}$->$\rho$!.
\end{lemma}
\begin{proof}
    By induction on derivation of !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$!.
\end{proof}

@@

We first give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}, which we call $\UNFSymbol$.
As discussed above, $\UNFSymbol$ sends a constant or an operator to a primitive in Source UNF. 
This term in Source UNF carries the context of the original term in Source.
$\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
Because of this, when we are trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
When we need to pass the result of type !A! of !e$_{1}$! through !e$_{2}$! which does not take it as an input, we can transform !e$_{2}$! to accept and pass it.
It's the equivalent in UNF of a weakening.
This weakening is required when we transform a non-unary operator such as $+$ or !pair!, as we transform the arguments in order and we need to keep all of their results.
We write this new term $(\widetilde{e_{2}})_{A}$ and it defined by induction on !e$_{2}$! as follows. 
We assume given a type !A! that needs to be passed, and drop the index !A!.

@@

Next, the transformation from Target UNF to Target is presented in Figure~\ref{fig:unf_to_target}. 
We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of the ANF transformation \cite{sabry1993reasoning}.

\begin{example}
    TODO
\end{example}

\begin{example}
    TODO
\end{example}

$\UNFSymbol^{-1}$ does not have as a simple typing property as $\UNFSymbol$, 
because it treats primitives with and without a $J$ differently and should be performed after $\Dsynrevsymbol_{\rho}$.
